\noindent
Owing to the complexity of the task, the handbook will not proceed in a 100\%
top-down fashion. Rather, we begin with a draft structure and a brief (one
page) synopsis for each chapter as planned {\it at a time when the chapter
structure is still fluid}. We begin by cannibalizing \cite{Kracht:2003}
and \cite{Kornai:2007}, which deal, roughly, with the discrete and the
continuous aspects of ML respectively. 

\subsection*{Volume 1: discrete}

We should have separate chapters on:

\begin{enumerate}
\item Formal languages
\item Generative capacity
\item Automata theory
\item Subregular
\item Finite state
\item Context free
\item Mildly CS
\item CS and beyond
\item Influential syntax formalisms: TG, LFG, GPSG, HPSG, MP, UD
\item Categorial Grammar, CCG
\item Model-theoretic syntax
\item Influential semantic formalisms
\item Knowledge Representation
\end{enumerate}

\subsection*{Volume 2: continuous}


\begin{enumerate}
\item Weighted generalizations of formal languages and automata
\item Phonetics
\item Speech synthesis
\item Phonology
\item Learnability
\item Speech Recognition
\item Optical character recognition
\item Information theory
\item Statistical linguistics
\item Historical linguistics
\item Psycholinguistics
\item Machine learning
\item Natural Language Processing
\item Machine Translation
\item Continuous vector semantics
\item Typology
\end{enumerate}


\noindent
The chapters (in both volumes) are given in no particular order. Some of them
already have candidate editors, most don't, so feel free to nominate yourself
by editing {\tt editors.tsv}, and adding a synopsis -- a mini-sample is
provided for 2.7.

Your chapter synopsis should not exceed 1 page, should cite some relevant
literature, and should be named subjectofchapter.tex. You should 
$\backslash$include it in {\tt synopsis.tex}. At this point, negative statements
(I'm planning on covering X and Y, but not Z) also make sense.

For ease of diff-ing, keep line length of .tex files to 80, and for ease of
merging bibliographies, name your entries Firstauthor:fullyear, and use
Smith:2017, Smith:2017a, Smith2017b if you need to cite three papers from
various people (or the same person) named Smith from the same year. Save us
the pain, avoid anything outside 7-bit ascii in the entry name (but feel free
to use utf8 in the author = field). Check ML/ml.bib and MoL/formal.bib to see
if your entry is already present before adding it to a new bib file (which
should be the same synopsis.bib for everyone). 


