\chapter*{Preface}
%
%
%
The present book developed out of lectures and seminars held over many 
years at the Department of Mathematics of the Freie Universit\"at Berlin, 
the Department of Linguistics of the Universit\"at Potsdam and the 
Department of Linguistics at UCLA. I wish to thank in particular 
the Department of Mathematics at the Freie Universit\"at Berlin 
as well as the Freie Universit\"at Berlin for their support and 
the always favourable conditions under which I was allowed to work.
Additionally, I thank the DFG for providing me with a 
Heisenberg--Stipendium, a grant that allowed me to continue 
this project in between various paid positions.

I have had the privilege of support by Hans--Martin G\"artner, 
%%%
\index{G\"artner, Hans--Martin}%%%
\index{Keenan, Edward L.}%%
\index{Kolb, Hap}%%%
\index{M\"onnich, Uwe}%%%
%%%
Ed Keenan, Hap Kolb and Uwe M\"onnich. Without them I would 
not have had the energy to pursue this work and fill so many 
pages with symbols that create so much headache. They always 
encouraged me to go on. 

Lumme Erilt, Greg Kobele and Jens Michaelis 
%%%
\index{Erilt, Lumme}%%%
\index{Kobele, Greg}%%%
\index{Michaelis, Jens}%%%
%%%
have given me invaluable help by scrupulously reading earlier 
versions of this manuscript. Further, I wish to thank Helmut Alt, 
Christian Ebert, 
%%%
\index{Alt, Helmut}%%%
\index{Ebert, Christian}%%%
\index{Fabian, Benjamin}%%%
\index{Gehrke, Stefanie}%%%
\index{Hanke, Timo}%%%
\index{Hodges, Wilfrid}%%%
\index{J\"ager, Gerhard}%%%
\index{Kanazawa, Makoto}%%%
\index{Koniecny, Franz}%%%
\index{Kosiol, Thomas}%%%
\index{Lin, Ying}%%%
\index{Lipt\'ak, Zsuzsanna}%%
\index{N\'emeti, Istv\'an}%%%
\index{Parsons, Terry}%%%
\index{Manaster--Ramer, Alexis}%%%
\index{Riggle, Jason}%%%
\index{Salinger, Stefan}%%%
\index{Stabler, Edward P.}%%%
\index{Sternefeld, Wolfgang}%%%
\index{Stamm, Harald}%%
\index{Staudacher, Peter}%%%
\index{Tchao, Ngassa}%%%
%%%
Benjamin Fabian, Stefanie Gehrke, Timo Hanke, Wilfrid Hodges, Gerhard 
J\"ager, Makoto Kanazawa, Franz Koniecny, Thomas Kosiol, Ying Lin, 
Zsu\-zsan\-na Lipt\'ak, Istv\'an N\'emeti, Terry Parsons, Alexis--Manaster 
Ramer, Jason Riggle, Stefan Salinger, Ed Stabler, Harald Stamm, Peter 
Staudacher, Wolfgang Sternefeld and Ngassa Tchao for their help.
%%%
\\[1cm]
Los Angeles and Berlin, September 2003
\hfill 
{\it Marcus Kracht}
