\newenvironment{eintrag}{\begin{tabular}{@{\quad}p{1.6cm}@{}p{9.8cm}}}%
{\end{tabular}}
\chapter*{Bibliography}
\markboth{Bibliography}{}
%%%
Ajdukiewicz, Kazimierz 
\\
\begin{eintrag}
1936 & Die syntaktische {K}onnexit\"at [The syntactic connectivity]. 
        {\em Studia Philosophica} 1:1 -- 27.
\end{eintrag}
\\[2.5mm]
Barendregt, Henk
\\
\begin{eintrag}
1985 & {\em The Lambda Calculus. Its Syntax and Semantics}.
	Studies in Logic and the Foundations of Mathematics 103. 
	2nd ed. Amsterdam: Elsevier.
\end{eintrag}
\\[2.5mm]
Barker, Chris, and Pullum, Geoffrey.
\\
\begin{eintrag}
1990 & A theory of command relations. {\em Linguistics and Philosophy} 
13:1--34.
\end{eintrag}
\\[2.5mm]
Bauer, Brigitte L.~M. 
\\
\begin{eintrag}
1995 & {\em The {E}mergence and {D}evelopment of {SVO} {P}atterning in
  {L}atin and {F}rench}. Oxford: Oxford University Press.
\end{eintrag}
\\[2.5mm]
Bird, Steven, and Ellison, Mark
\\
\begin{eintrag}
1994 & One--level phonology: {A}utosegmental representations and rules as
  finite automata. {\em Computational Linguistics} 20:55 -- 90.
\end{eintrag}
\\[2.5mm]
Blackburn, Patrick
\\
\begin{eintrag}
1993 & Modal logic and attribute value structures.  In {\em Diamonds and 
	Defaults}, Maarten de Rijke (ed.),  19 -- 65.  (Synthese
	Library 229.) Dordrecht: Kluwer.
\end{eintrag}
\\[2.5mm]
Blok, Wim J., and Pigozzi, Don~J. 
\\
\begin{eintrag}
1990 & Algebraizable logics. 
{\em Memoirs of the Americal Mathematical Society}, 77(396).
\end{eintrag}
\\[2.5mm]
Bochvar, D.~A.
\\
\begin{eintrag}
1938 & On a three--valued logical calculus and its application to the
  analysis of contradictions.
  {\em Mathematicheskii Sbornik} 4:287 -- 308.
\end{eintrag}
\\[2.5mm]
B\"ottner, Michael, and Th\"ummel, Wolf (eds.)
\\
\begin{eintrag}
2000 & {\em Variable--free {S}emantics}. Artikulation und
  Sprache 3. Osnabr\"uck: secolo Verlag.
\end{eintrag}
\\[2.5mm]
Bresnan, Joan, Kaplan, Ronald~M., Peters, Stanley, and Zaenen, Annie
\\
\begin{eintrag}
1987 & Cross--{S}erial {D}ependencies in {D}utch. In {\em The {F}ormal 
  {C}omplexity of {N}atural {L}anguage}, Walter Savitch, Emmon Bach, William 
  Marsch, and Gila Safran--Naveh (eds.), 286 -- 319. Dordrecht: Reidel.
\end{eintrag}
\\[2.5mm]
B\"uchi, J.
\\
\begin{eintrag}
1960 & Weak second--order arithmetic and finite automata.
  {\em Zeitschrift f\"ur Mathematische Logik und Grundlagen der
  Mathematik} 6:66 -- 92.
\end{eintrag}
\newpage
\noindent
Burmeister, Peter
\\
\begin{eintrag}
1986 &  {\em A {M}odel {T}heoretic {O}riented {A}pproach to {P}artial
  {A}lgebras}. Berlin: Akademie Verlag. \\
2002 & {L}ecture {N}otes on {U}niversal {A}lgebra. {M}any {S}orted {P}artial
  {A}lgebras. Manuscript available via internet.
\end{eintrag}
\\[3mm]
Burris, Stanley, and Sankappanavar, H.~P.
\\
\begin{eintrag}
1981 & {\em A Course in Universal Algebra}.  Graduate Texts in 
	Mathematics 78. Berlin/New York: Springer.
\end{eintrag}
\\[3mm]
Buszkowski, Wojciech
\\
\begin{eintrag}
1997 & Mathematical linguistics and proof theory.
   In {\em Handbook of Logic and Language}, Johan van Benthem and 
   Alice ter Meulen (eds.), 683 -- 736. Amsterdam: Elsevier.
\end{eintrag}
\\[3mm]
Carpenter, Bob
\\
\begin{eintrag}
1992 & {\em The Logic of Typed Feature Structures}.
	Cambridge Tracts in Theoretical Computer Science 32. Cambridge: 
	Cambridge University Press.
\end{eintrag}
\\[3mm]
Chandra, A.~K., Kozen, D.~C., and Stockmeyer, L.~J.
\\
\begin{eintrag}
1981 & Alternation. {\em Journal of the Association for Computing Machinery} 
	28:114 -- 133.
\end{eintrag}
\\[3mm]
Chomsky, Noam and Halle, Morris
\\
\begin{eintrag}
1968 & {\em The sound pattern of {E}nglish}.
       New York: Harper and Row.
\end{eintrag}
\\[3mm]
Chomsky, Noam
\\
\begin{eintrag}
1959 & On certain formal properties of grammars.
       {\em Information and Control} 2:137 -- 167.
\\
1962 & Context--free grammars and pushdown storage.  {\em MIT Research 
  Laboratory of Electronics Quarterly Progress Report} 65.
\\
1981 & {\em Lecture Notes on Government and Binding}. Dordrecht: Foris.
\\
1986 & {\em Barriers}. Cambridge (Mass.): MIT Press.
\\
1993 & A minimalist program for linguistic theory. In {\em The View from 
	Building 20: Essays in Honour of Sylvain Bromberger}, Ken Hale and 
	Samuel J. Keyser (eds.), 1 -- 52. Cambridge (Mass.): MIT Press.
\end{eintrag}
\\[3mm]
Church,  Alonzo
\\
\begin{eintrag}
1933 & A set of postulates for the foundation of logic.
       {\em Annals of Mathematics} 2:346 -- 366.
\\
1940 & A formulation of the simple theory of types.
       {\em Journal of Symbolic Logic} 5:56 -- 68.
\end{eintrag}
\\[3mm]
Coulmas, Florian
\\
\begin{eintrag}
2003 & {\em Writing Systems. An introduction to their linguistic 
	analysis}. Cambridge: Cambridge University Press.
\end{eintrag}
\newpage
\noindent
Culy, Christopher 
\\
\begin{eintrag}
1987 & The {C}omplexity of the {V}ocabulary of {B}ambara.  In 
	{\em The {F}ormal {C}omplexity of {N}atural {L}anguage}, 
	Walter Savitch, Emmon Bach, William Marsch, and Gila Safran--Naveh
        (eds.), 345 -- 351. Dordrecht: Reidel.
\end{eintrag}
\\[2.8mm]
Curry, Haskell~B. 
\\
\begin{eintrag}
1930 & Grundlagen der kombinatorischen {L}ogik [{F}oundations of
  combinatory logic]. {\em American Journal of Mathematics} 
  52:509 -- 536, 789 -- 834.
\\
1977 & {\em Foundations of Mathematical Logic}. 2nd ed. New York: 
	Dover Publications.
\end{eintrag}
\\[2.8mm]
Davey, B.~A., and Priestley, H.~A.
\\
\begin{eintrag}
1990 & {\em Lattices and Order}. Cambridge: Cambridge University Press.
\end{eintrag}
\\[2.8mm]
Deutsch, David, Ekert, Artur, and Luppacchini, Rossella
\\
\begin{eintrag}
2000 & Machines, logic and quantum physics. {\em Bulletin of Symbolic 
	Logic} 6:265 -- 283.
\end{eintrag}
\\[2.8mm]
Doner, J.~E.
\\
\begin{eintrag}
1970 & Tree acceptors and some of their applications.
       {\em Journal of Computer and Systems Sciences} 4:406 -- 451.
\end{eintrag}
\\[2.8mm]
Dowty, David~R., Wall, Robert~E., and Peters, Stanley 
\\
\begin{eintrag}
1981 & {\em Introduction to Montague Semantics}. Synthese Library 11. 
	Dordrecht: Reidel.
\end{eintrag}
\\[2.8mm]
Dresner, Eli
\\
\begin{eintrag}
2001 & Tarski's {R}estricted {F}orm and {N}eale's {Q}uantificational
  {T}reatment of {P}roper {N}ames. {\em Linguistics and Philosophy} 
  24:405 -- 415.
\\
2002 & Holism, {L}anguage {A}cquisition, and {A}lgebraic {L}ogic.
     {\em Linguistics and Philosophy} 25:419 -- 452.
\end{eintrag}
\\[2.8mm]
Dymetman, Marc
\\
\begin{eintrag}
1991 & Inherently reversible grammars, logic programming and computability.
   In {\em Proceedings of the ACL Workshop: Reversible Grammars in
   Natural Language Processing}.
\\
1992 & {\em Transformations de Grammaires Logiques et R\'eversibilit\'e}
  [Transformations of Logical Grammars and Reversibility].
   Ph.\ D.\  diss., Universit\'e Joseph Fourier, Grenoble.
\end{eintrag}
\\[2.8mm]
Ebbinghaus, Hans-Dieter, and Flum, J\"org 
\\
\begin{eintrag}
1995 & {\em Finite {M}odel {T}heory}. Perspectives in Mathematical 
	Logic. Berlin/New York: Springer.
\end{eintrag}
\\[2.8mm]
Ebert, Christian, and Kracht, Marcus
\\\begin{eintrag}
2000 & Formal syntax and semantics of case stacking languages.
        In {\em Proceedings of the EACL 2000}.
\end{eintrag}
\newpage
\noindent
van Eijck, Jan
\\\begin{eintrag}
1994 & Presupposition failure: a comedy of errors. {\em Formal Aspects 
	of Computing} 3.
\end{eintrag}
\\[1.95mm]
Eisenberg, Peter
\\\begin{eintrag}
1973 & A {N}ote on `{I}dentity of {C}onstituents'. {\em Linguistic 
	Inquiry} 4:417 -- 420.
\end{eintrag}
\\[1.95mm]
Ewen, Colin~J., and van~der Hulst, Harry
\\\begin{eintrag}
2001 & {\em The Phonological Structure of Words}.
	Cambridge: Cambridge University Press.
\end{eintrag}
\\[1.95mm]
Ferreir\'os, Jos\'e
\\\begin{eintrag}
2001 & The road to modern logic --- an interpretation.
    {\em The Bulletin of Symbolic Logic} 7:441 -- 484.
\end{eintrag}
\\[1.95mm]
Fiengo, Robert, and May, Robert 
\\
\begin{eintrag}
1994 & {\em Indices and Identity}. Linguistic Inquiry Monographs 24. 
	Cambridge (Mass.): MIT Press.
\end{eintrag}
\\[1.95mm]
Fine, Kit
\\\begin{eintrag}
1992 & Transparency, {P}art {I}: {R}eduction. Unpublished manuscript, UCLA.
\end{eintrag}
\\[1.95mm]
Frege, Gottlob 
\\\begin{eintrag}
1962 & Funktion und {B}egriff [{F}unction and {C}oncept].
  In {\em Funktion, Begriff, Bedeutung.  F\"unf logische Studien} 
	[{F}unction, {C}oncept, {M}eaning. {F}ive logical {S}tudies], 
	G\"unther Patzig (ed.), 17 -- 39. G\"ottingen: Vandenhoeck \& Ruprecht.
\end{eintrag}
\\[1.95mm]
Frey, Werner 
\\\begin{eintrag}
1993 & {\em {S}yntaktische {B}edingungen f\"{u}r die semantische
  {I}nterpretation} [Syntactic {C}onditions for the {S}emantic 
  {I}nterpretation]. Number~35 in Studia Grammatica. Berlin: 
  Akademie Verlag.
\end{eintrag}
\\[1.95mm]
Fromkin, V. (ed.)
\\\begin{eintrag}
2000 & {\em Linguistics: An Introduction to linguistic theory}.
         London: Blackwell.
\end{eintrag}
\\[1.95mm]
Gamut, L.~T.~F. 
\\\begin{eintrag}
1991a & {\em Logic, Language and Meaning. Vol. 1: Introduction to 
	Logic}. Chicago: The University of Chicago Press.
\\
1991b & {\em Logic, Language and Meaning. Vol. 2: Intensional Logic and
  Logical Grammar}. Chicago: The University of Chicago Press.
\end{eintrag}
\\[1.95mm]
G\"ardenfors, Peter
\\\begin{eintrag}
1988 & {\em Knowledge in Flux}. Cambridge (Mass.): MIT Press.
\end{eintrag}
\\[1.95mm]
Gazdar, Gerald, Klein, Ewan, Pullum, Geoffrey, and Sag, Ivan.
\\\begin{eintrag}
1985 & {\em Generalized Phrase Structure Grammar}. London: Blackwell.
\end{eintrag}
\\[2mm]
Gazdar, Gerald, Pullum, Geoffrey, Carpenter, Bob, Hukari, T., and Levine, R.
\\\begin{eintrag}
1988 & Category structures. {\em Computational Linguistics} 14:1 -- 19.
\end{eintrag}
\newpage
\noindent
Geach, Peter 
\\\begin{eintrag}
1972 & A {P}rogram for {S}yntax. In {\em Semantics for Natural Language}, 
	Donald Davidson and Gilbert Harman (eds.). (Synthese Library 40.) 
	Dordrecht: Reidel.
\end{eintrag}
\\[2.8mm]
Geller, M.~M., and  Harrison, M.~A.
\\\begin{eintrag}
1977 & On {$LR(k)$} grammars and languages. {\em Theoretical Computer 
	Science} 4:245 -- 276.
\end{eintrag}
\\[2.8mm]
Geurts, Bart
\\\begin{eintrag}
1998 & Presupposition and {A}naphors in {A}ttitude {C}ontexts.
    {\em Linguistics and Philosophy} 21:545 -- 601.
\end{eintrag}
\\[2.8mm]
Ginsburg, Seymour and Spanier, Edwin~H. 
\\\begin{eintrag}
1964 & Bounded {ALGOL}--{L}ike {L}anguages. {\em Transactions of the 
	American Mathematical Society} 113:333 -- 368.
\\
1966 & Semigroups, {P}resburger {F}ormulas, and {L}anguages.
	{\em Pacific Journal of Mathematics} 16:285 -- 296.
\end{eintrag}
\\[2.8mm]
Ginsburg, Seymour
\\\begin{eintrag}
1975 & {\em Algebraic and Automata--Theoretic Properties of Formal
  Languages}. Amsterdam: North--Holland.
\end{eintrag}
\\[2.8mm]
Goldstern, Martin, and Judah, Haim 
\\\begin{eintrag}
1995 & {\em The Incompleteness Phenomenon}. Wellesley (Mass.): 
	AK Peters.
\end{eintrag}
\\[2.8mm]
Gr\"atzer, George 
\\\begin{eintrag}
1968 & {\em Universal Algebra}. New York: van Nostrand.
\\
1971 & {\em Lattice {T}heory: {F}irst {C}oncepts and {D}istributive
  {L}attices}. Freeman.
\end{eintrag}
\\[2.8mm]
Greibach, Sheila~A. 
\\\begin{eintrag}
1967 & A new normal--form theorem for context--free phrase structure
  grammars. {\em Journal of the Association for Computing Machinery} 
  13:42 -- 52.
\end{eintrag}
\\[2.8mm]
Grewendorf, G\"unter, Hamm, Friedrich, and Sternefeld, Wolfgang
\\\begin{eintrag}
1987 & {\em Sprachliches Wissen. Eine Einf\"uhrung in moderne Theorien der
  grammatischen Beschreibung\/} [{K}nowledge of {L}anguage. 
	An {I}ntroduction to {M}odern {T}heories of {G}rammatical 
	{D}escription].
  Number 695 in suhrkamp taschenbuch wissenschaft. Frankfurt a.M.: 
	Suhrkamp Verlag.
\end{eintrag}
\\[2.8mm]
Groenink, Annius~V. 
\\\begin{eintrag}
1997a & Mild context--sensitivity and tuple--based generalizations of
  context--grammar. {\em Linguistics and Philosophy} 20:607--636.
\\
1997b & {\em Surface without Structure. Word Order and Tractability Issues in
  Natural Language Ana\-ly\-sis}. Ph.\ D.\ diss., University of Utrecht.
\end{eintrag}
\newpage\noindent
de~Groote, Philippe
\\
\begin{eintrag}
2001 & Towards {A}bstract {C}ategorial {G}rammars.
  In {\em Association for Computational Linguistics, 39th Annual
  Meeting and 10th Conference of the European Chapter}, 148 -- 155,
  Toulouse.
\end{eintrag}
\\[2.25mm]
Haider, Hubert 
\\\begin{eintrag}
1991 & Die menschliche Sprachf\"ahigkeit --- exaptiv und kognitiv opak
	[The human language faculty --- exaptive and cognitively opaque].
       {\em Kognitionswissenschaft} 2:11 -- 26.
\\
1993 & {\em Deutsche Syntax --- generativ. Vorstudien zur Theorie einer
       projektiven Grammatik} [German Syntax --- generative. Preliminary 
       studies towards a theory of projective grammar]. T\"ubingen: 
       Gunter Narr Verlag.
\\
1995 & Downright down to the right. In {\em On {E}xtraction and 
	{E}xtraposition}, U.~Lutz and J.~Pafel (eds.), 145 -- 271. 
	Amsterdam: John Benjamins.
\\
1997 & Extraposition. In {\em Rightward Movement}, 
	D.~Beerman, D.~LeBlanc, and H.~van Riemsdijk (eds.), 
	115 -- 151. Amsterdam: John Benjamins.
\\
2000 & {B}ranching and {D}ischarge. In {\em Lexical Specification and 
	Insertion}, Peter Coopmans, Martin Everaert, and Jane Grimshaw 
	(eds.), 135 -- 164. (Current Issues in Linguistic Theory 197.) 
        Amsterdam: John Benjamins.
\end{eintrag}
\\[2.25mm]
Halmos, Paul 
\\\begin{eintrag}
1956 & Homogeneous locally finite polyadic boolean algebras of infinite
  degree. {\em Fundamenta Mathematicae} 43:255 -- 325.
\end{eintrag}
\\[2.25mm]
Hamm, Friedrich, and van Lambalgen, Michiel 
\\\begin{eintrag}
2003 & Event Calculus, Nominalization and the Progressive. {\em Linguistics 
and Philosophy} 26:381 -- 458.
\end{eintrag}
\\[2.25mm]
Harkema, Henk
\\\begin{eintrag}
2001 & A {C}haracterization of {M}inimalist {L}anguages.  In 
        {\em Logical Aspects of Computational Linguistics (LACL '01)}, 
	Philippe de~Groote, Glyn Morrill, and Christian Retor\'e (eds.),
	193 -- 211. (Lecture Notes in Artificial Intelligence 2099.)
	Berlin/New York: Springer.
\end{eintrag}
\\[2.25mm]
Harris, Zellig~S. 
\\\begin{eintrag}
1963 & {\em Structural Linguistics}. The University of Chicago Press.
\\
1979 & {\em Mathematical Structures of Language}. Huntington (NY): 
	Robert E. Krieger Publishing Company.
\end{eintrag}
\\[2.25mm]
Harrison, Michael~A. 
\\\begin{eintrag}
1978 & {\em Introduction to Formal Language Theory}. 
	Reading (Mass.): Addison Wesley.
\end{eintrag}
\\[2.25mm]
Hausser, Roland~R. 
\\\begin{eintrag}
1984 & {\em Surface Compositional Grammar}. M\"unchen: Wilhelm Finck 
	Verlag.
\end{eintrag}
\newpage\noindent
Heim, Irene 
\\\begin{eintrag}
1983 & On the projection problem for presuppositions. In 
	{\em Proceedings of the 2nd West Coast Conference on Formal 
	Linguistics}, M.~Barlow and D.~Flickinger, D.~Westcoat (eds.), 
        114 -- 126, Stanford University.
\end{eintrag}
\\[1.9mm]
Hendriks, Herman 
\\\begin{eintrag}
2001 & Compositionality and {M}odel--{T}heoretic {I}nterpretation.
{\em Journal of Logic, Language and Information} 10:29 -- 48.
\end{eintrag}
\\[1.9mm]
Henkin, Leon, Monk, Donald, and Tarski, Alfred 
\\\begin{eintrag}
1971 & {\em Cylindric Algebras. Part 1}. Studies in Logic 
	and the Foundation of Mathematics 64.  Amsterdam: North--Holland.
\end{eintrag}
\\[1.9mm]
Hindley, J.~R., Lercher, B., and Seldin, J.~P.
\\\begin{eintrag}
1972 & {\em Introduction to Combinatory Logic}. London Mathematical 
	Society Lecture Notes 7. Oxford: Oxford University Press.
\end{eintrag}
\\[1.9mm]
Hindley, J.~R. and Longo, L.
\\\begin{eintrag}
1980 & Lambda calculus models and extensionality. {\em Zeitschrift 
	f\"ur mathematische Logik und Grundlagen der Mathematik} 
	26:289 -- 310.
\end{eintrag}
\\[1.9mm]
Hintikka, Jaakko
\\\begin{eintrag}
1962 & {\em Knowledge and Belief. An Introduction into the logic of 
	the two notions}. Ithaca: Cornell University Press.
\end{eintrag}
\\[1.9mm]
Hodges, Wilfrid 
\\\begin{eintrag}
2001 & Formal features of compositionality. {\em Journal of Logic, 
	Language and Information} 10:7 -- 28.
\end{eintrag}
\\[1.9mm]
Hopcroft, John~E., and Ullman, Jeffrey~D. 
\\\begin{eintrag}
1969 & {\em Formal Languages and their Relation to Automata}.
	Reading (Mass.): Addison Wesley.
\end{eintrag}
\\[1.9mm]
van der Hulst, Harry
\\\begin{eintrag}
1984 & {\em Syllable Structure and Stress in Dutch}. Dordrecht: Foris.
\end{eintrag}
\\[1.9mm]
Huybregts, Riny
\\\begin{eintrag}
1984 & {O}verlapping {D}ependencies in {D}utch. {\em Utrecht Working 
	Papers in Linguistics} 1:3 -- 40.
\end{eintrag}
\\[1.9mm]
IPA
\\\begin{eintrag}
1999 & {\em Handbook of the International Phonetic Association}.
	Cambridge: Cambridge University Press.
\end{eintrag} 
\\[1.9mm]
Jackendoff, Ray
\\\begin{eintrag}
1977 & {\em $\oli{X}$--{S}yntax: {A} {S}tudy of {P}hrase {S}tructure}.
	Linguistic Inquiry Monographs 2. Cambridge (Mass.): 
	MIT Press.
\end{eintrag}
\\[1.9mm]
Jacobson, Pauline 
\\\begin{eintrag}
1999 & Toward a {V}ariable--{F}ree semantics. {\em Linguistics and 
	Philosophy} 22:117 -- 184.
\end{eintrag}
\newpage\noindent
\begin{eintrag}
2002 & The ({D}is){O}rganisation of the {G}rammar: 25 {Y}ears.
	{\em Linguistics and Philosophy} 25:601 -- 626.
\end{eintrag}
\\[3.4mm]
Johnson, J.~S.
\\\begin{eintrag}
1969 & Nonfinitizability of classes of representable polyadic algebras.
	{\em Journal of Symbolic Logic} 34:344 -- 352.
\end{eintrag}
\\[3.4mm]
Johnson, Mark
\\\begin{eintrag}
1988 & {\em Attribute--Value Logic and the Theory of Grammar}. 
	CSLI Lecture Notes 16. Stanford: CSLI.
\end{eintrag}
\\[3.4mm]
Jones, Burton~W. 
\\\begin{eintrag}
1955 & {\em The Theory of Numbers}. New York: Holt, Rinehart and Winston. 
\end{eintrag}
\\[3.4mm]
Joshi, Aravind, Levy, Leon~S., and Takahashi, Masako 
\\\begin{eintrag}
1975 & Tree {A}djunct {G}rammars. {\em Journal of Computer and System 
	Sciences} 10:136 -- 163.
\end{eintrag}
\\[3.4mm]
Joshi, Aravind~K. 
\\\begin{eintrag}
1985 & Tree adjoining grammars: {H}ow much context--sensitivity is 
	required to provide reasonable structural descriptions?
	In {\em Natural {L}anguage {P}arsing. {P}sychological, 
	{C}omputational, and {T}heoretical {P}erspectives}, 
        David Dowty, Lauri Karttunen, and Arnold Zwicky (eds.), 
	206--250. Cambridge: Cambridge University Press.
\end{eintrag}
\\[3.4mm]
Just, Winfried and Weese, Martin 
\\\begin{eintrag}
1996 & {\em Discovering Modern Set Theory. Vol.\ I: The Basics}. 
	Graduate Studies in Mathematics 8. AMS.
\\
1997 & {\em Discovering Modern Set Theory. Vol.\ II: Set--Theoretic 
	Tools for Every Mathematician}. Graduate Studies in Mathematics 
	18.  AMS.
\end{eintrag}
\\[3.4mm]
Kac, Michael~B., Manaster--Ramer, Alexis, and Rounds, William~C. 
\\\begin{eintrag}
1987 & Simultaneous--{D}istributive {C}oordination and {C}ontext--{F}reeness.
	{\em Computational Linguistics} 13:25 -- 30.
\end{eintrag}
\\[3.4mm]
Kaplan, Ron~M., and Kay, Martin 
\\\begin{eintrag}
1994 & Regular {M}odels of {P}honological {R}ule {S}ystems.
	{\em Computational Linguistics} 20:331 -- 378.
\end{eintrag}
\\[3.4mm]
Karttunen, Lauri
\\\begin{eintrag}
1974 & Presuppositions and linguistic context. {\em Theoretical Linguistics} 
	1:181 -- 194.
\end{eintrag}
\\[3.4mm]
Kasami, Tadao, Seki, Hiroyuki, and Fujii, Mamoru 
\\\begin{eintrag}
1987 & Generalized context--free grammars, multiple context--free 
	grammars and head grammars. Technical report, Osaka University.
\end{eintrag}
%\\[2mm]
\newpage\noindent
Kasami, Tadao 
\\\begin{eintrag}
1965 & An efficient recognition and syntax--analysis algorithm for
	context--free languages. Technical report, Air Force Cambridge 
	Research Laboratory, Bedford (Mass.). Science Report AFCRL--65--758.
\end{eintrag}
\\[2.1mm]
Kayne, Richard~S. 
\\\begin{eintrag}
1994 & {\em The {A}ntisymmetry of {S}yntax}. Linguistic 
	Inquiry Monographs 25. Cambridge (Mass.): MIT Press.
\end{eintrag}
\\[2.1mm]
Keenan, Edward~L., and Faltz, Leonard~L. 
\\\begin{eintrag}
1985 & {\em Boolean Semantics for Natural Language}. Dordrecht: Reidel.
\end{eintrag}
\\[2.1mm]
Keenan, Edward L., and Westerst{\aa}hl, Dag
\\\begin{eintrag}
1997 & Generalized quantifiers. In {\em Handbook of Logic and Language}, 
	Johan van Benthem and Alice ter Meulen (eds.), 835 -- 893. 
	Amsterdam: Elsevier.
\end{eintrag}
\\[2.1mm]
Kempson, Ruth
\\\begin{eintrag}
1975 & {\em Presupposition and the delimitation of semantics}. 
	Cambridge: Cambridge University Press.
\end{eintrag}
\\[2.1mm]
Kleene, Stephen~C. 
\\\begin{eintrag}
1956 & Representation of events in nerve nets. In {\em Automata Studies}, 
	C.~E. Shannon and J.~McCarthy (eds.), 3 -- 40. 
	Princeton: Princeton University Press.
\end{eintrag}
\\[2.1mm]
Knuth, Donald 
\\\begin{eintrag}
1956 & On the translation of languages from left to right.
       {\em Information and Control} 8:607 -- 639.
\end{eintrag}
\\[2.1mm]
Koskenniemi, Kimmo
\\\begin{eintrag}
1983 & Two--level morphology. {A} general computational model for 
	word--form recognition. Technical Report~11, Department of 
	General Linguistics, University of Helsinki.
\end{eintrag}
\\[2.1mm]
Koster, Jan
\\\begin{eintrag}
1986 & {\em Domains and Dynasties: {T}he Radical Autonomy of Syntax}.
	Dordrecht: Foris.
\end{eintrag}
\\[2.1mm]
Koymans, J.~P.~C. 
\\\begin{eintrag}
1982 & Models of the {L}ambda {C}alculus. {\em Information and 
	Control} 52:306 -- 332.
\end{eintrag}
\\[2.1mm]
Kracht, Marcus
\\\begin{eintrag}
1993 & Mathematical aspects of command relations. In {\em Proceedings 
	of the EACL 93}, 241 -- 250.
\\
1994 & Logic and {C}ontrol: {H}ow {T}hey {D}etermine the {B}ehaviour of
	{P}resuppositions. In {\em Logic and Information Flow}, 
	Jan van Eijck and Albert Visser (eds.), 88 -- 111. 
	Cambridge (Mass.): MIT Press.
\end{eintrag}
\newpage\noindent
\begin{eintrag}
1995a & Is there a genuine modal perspective on feature structures?
	{\em Linguistics and Philosophy} 18:401 -- 458. \\
1995b & Syntactic {C}odes and {G}rammar {R}efinement. {\em Journal 
	of Logic, Language and Information} 4:41 -- 60.
\\
1997 & Inessential {F}eatures. In {\em Logical Aspects of Computational 
	Linguistics (LACL '96)}, Christian Retor\'{e} (ed.), 43 -- 62. 
	(Lecture Notes in Artificial Intelligence 1328.)
	Berlin/New York: Springer.
\\
1998 & Adjunction {S}tructures and {S}yntactic {D}omains. In {\em The 
	Mathematics of Sentence Structure. Trees and Their Logics}, 
	Uwe M\"onnich and Hans-Peter Kolb (eds.), 259 -- 299. (Studies 
	in Generative Grammar 44.) Berlin: Mouton de Gruyter.
\\
1999 & {\em Tools and {T}echniques in {M}odal {L}ogic}. 
	Studies in Logic and the Foundations of Mathematics 142. 
	Amsterdam: Elsevier.
\\
2001a & Modal {L}ogics {T}hat {N}eed {V}ery {L}arge {F}rames.
	{\em Notre Dame Journal of Formal Logic} 42:141 -- 173.
\\
2001b & Syntax in {C}hains. {\em Linguistics and Philosophy}
	24:467 -- 529.
\\
2003 & Against the feature bundle theory of case. In 
	{\em New Perspectives on Case Theory}, Eleonore Brandner 
	and Heike Zinsmeister (eds.), 165 -- 190. Stanford: CSLI.
\end{eintrag}
\\[2.2mm]
Kuroda, S.~Y.
\\\begin{eintrag}
1964 & Classes of languages and linear bounded automata. {\em Information 
	and Control} 7:207 -- 223.
\end{eintrag}
\\[2.2mm]
Lamb, Sydney~M. 
\\\begin{eintrag}
1966 & {\em Outline of {S}tratificational {G}rammar}. Washington: 
	Georgetown University Press.
\end{eintrag}
\\[2.2mm]
Lambek, Joachim
\\\begin{eintrag}
1958 & The {M}athematics of {S}entence {S}tructure. 
	{\em The American Mathematical Monthly} 65:154 -- 169.
\end{eintrag}
\\[2.2mm]
Landweber, Peter~S. 
\\\begin{eintrag}
1963 & Three theorems on phrase structure grammars of type 1.
	{\em Information and Control} 6:131 -- 137.
\end{eintrag}
\\[2.2mm]
Langholm, Tore 
\\\begin{eintrag}
2001 & A {D}escriptive {C}haracterisation of {I}ndexed {G}rammars.
	{\em Grammars} 4:205 -- 262.
\end{eintrag}
\\[2.2mm]
Lehmann, Winfred~P. 
\\\begin{eintrag}
1993 & {\em Theoretical {B}ases of {I}ndo--{E}uropean {L}inguistics}.
	London: Routledge.
\end{eintrag}
\\[2.2mm]
Leibniz, Gottfried~Wilhelm 
\\\begin{eintrag}
2000 & {\em Die Grund\-la\-gen des lo\-gi\-schen Kal\-k\"uls
        (Lateinisch--Deutsch)} [The {F}oundations of the Logical Calculus
	(Latin--German)]. Phi\-lo\-so\-phi\-sche Bibliothek 525. 
	Hamburg: Meiner Verlag.
\end{eintrag}
\\[2mm]
% Seitenumbruch
Levelt, Willem~P. 
\\\begin{eintrag}
1991 & {\em Speaking. From Intention to Articulation}. 2nd ed.
	Cambridge (Mass.): MIT Press.
\end{eintrag}
\\[3.2mm]
Lyons, John 
\\\begin{eintrag}
1968 & {\em Introduction to Theoretical Linguistics}. Cambridge: 
	Cambridge University Press.
\\
1978 & {\em Semantics. Vol. 1}. Cambridge: Cambridge University Press. 
\end{eintrag}
\\[3.2mm]
Manaster--Ramer, Alexis, and Kac, Michael~B. 
\\\begin{eintrag}
1990 & The {C}oncept of {P}hrase {S}tructure. {\em Linguistics and 
	Philosophy} 13:325 -- 362.
\end{eintrag}
\\[3.2mm]
Manaster--Ramer, Alexis, Moshier, M.~Andrew, and Zeitman, R.~Suzanne 
\\\begin{eintrag}
1992 & An {E}xtension of {O}gden's {L}emma. Manuscript. Wayne State 
	University, 1992.
\end{eintrag}
\\[3.2mm]
Manaster--Ramer, Alexis
\\\begin{eintrag}
1986 & 
Copying in natural languages, context--freeness and queue grammars.
In {\em Proceedings of the 24th Annual Meeting of the Association for
  Computational Linguistics}, 85 -- 89.
\end{eintrag}
\\[3.2mm]
Markov, A.~A.
\\\begin{eintrag}
1947 & On the impossibility of certain algorithms in the theory of
  associative systems ({R}ussian). {\em Doklady Akad\'emii Nauk SSSR} 
	55:587 -- 590.
\end{eintrag}
\\[3.2mm]
Marsh, William, and Partee, Barbara~H. 
\\\begin{eintrag}
1987 & How {N}on--{C}ontext {F}ree is {V}ariable {B}inding?
   	In {\em The {F}ormal {C}omplexity of {N}atural {L}anguage}, 
	Walter Savitch, Emmon Bach, William Marsch, and Gila 
	Safran--Naveh (eds.), 369 -- 386. Dordrecht: Reidel.
\end{eintrag}
\\[3.2mm]
Mel'\v{c}uk, Igor
\\\begin{eintrag}
1988 & {\em Dependency Syntax: Theory and Practice}. SUNY Linguistics 
	Series. Albany: State University of New York Press.
\\
2000 & {\em Cours de Morphologie G\'e\-n\'e\-rale\/} [General Morphology. A
 	 Coursebook]. Volume 1 -- 5. Montr\'eal: Les Presses de 
	l'Universit\'e de Montr\'eal.
\end{eintrag}
\\[3.2mm]
Meyer, A.~R.
\\\begin{eintrag}
1982 & What is a model of the lambda calculus? {\em Information and 
	Control} 52:87 -- 122.
\end{eintrag}
\\[3.2mm]
Michaelis, Jens, and Kracht, Marcus 
\\\begin{eintrag}
1997 & Semilinearity as a syntactic invariant. In {\em Logical Aspects 
	of Computational Linguistics (LACL '96)}, Christian Retor\'e (ed.), 
 	329 -- 345. (Lecture Notes in Artificial Intelligence 1328.)
	Heidelberg: Springer.
\end{eintrag}
\newpage\noindent
Michaelis, Jens, and Wartena, Christian 
\\\begin{eintrag}
1997 & How linguistic constraints on movement conspire to yield 
	languages analyzable with a restricted form of {LIGs}.
	In {\em Proceedings of the Conference on Formal Grammar 
	(FG '97), \emph{Aix en Provence}}, 158--168.
\\
1999 & {LIG}s with reduced derivation sets. In {\em Constraints and 
	Resources in Natural Language Syntax and Semantics}, Gosse Bouma, 
	Geert{--}Jan~M. Kruijff, Erhard Hinrichs, and Richard~T. 
	Oehrle (eds.), 263--279. Stanford: CSLI.
\end{eintrag}
\\[4.6mm]
Michaelis, Jens
\\\begin{eintrag}
2001a & Derivational minimalism is mildly context--sensitive.
	In {\em Logical Aspects of Computational Linguistics (LACL '98)},
	Michael Moortgat (ed.), 179 -- 198. (Lecture Notes in Artificial 
	Intelligence 2014.) Heidelberg: Springer.
\\
2001b & {\em On {F}ormal {P}roperties of {M}inimalist {G}rammars}.
	Ph.\ D.\ diss., Universit\"at Potsdam.
\\
2001c & Transforming linear context--free rewriting systems into 
	minimalist grammars. In {\em Logical Aspects of Computational 
	Linguistics (LACL '01)}, Philippe de~Groote, Glyn Morrill, and 
	Christian Retor{\'e} (eds.), 228 -- 244. (Lecture Notes 
	in Artificial Intelligence 2099.) Heidelberg: Springer.
\end{eintrag}
\\[4.6mm]
Miller, Philip~H. 
\\\begin{eintrag}
1991 & {S}candinavian {E}xtraction {P}henomena {R}evisited: {W}eak and
  {S}trong {G}enerative {C}apacity. {\em Linguistics and Philosophy} 
	14:101 -- 113.
\\
1999 & {\em Strong Generative Capacity. The Semantics of Linguistic
  Formalisms}. Stanford: CSLI.
\end{eintrag}
\\[4.6mm]
Mitchell, J.~C. 
\\\begin{eintrag}
1990 & Type systems for programming languages. In {\em Handbook of 
	Theoretical Computer Science, Vol B. Formal Models and Semantics}, 
	Jan van Leeuwen (ed.), 365 -- 458. Amsterdam: Elsevier.
\end{eintrag}
\\[4.6mm]
Monk, Donald~J. 
\\\begin{eintrag}
1969 & Nonfinitizability of classes of representable cylindric 
	algebras. {\em Journal of Symbolic Logic} 34:331 -- 343.
\\
1976 & {\em Mathematical Logic}. Berlin, Heidelberg: Springer.
\end{eintrag}
\\[4.6mm]
M\"onnich, Uwe
\\\begin{eintrag}
1999 & On cloning context--freeness. In {\em The Mathematics of Sentence 
	Structure. Trees and their Logics}, Hans-Peter Kolb and Uwe 
	M\"onnich (eds.), 195 -- 229. (Studies in Generative Grammar 44.) 
	Berlin: Mouton de Gruyter.
\end{eintrag}
\newpage\noindent
Moschovakis, Yannis
\\\begin{eintrag}
1994 & Sense and denotation as algorithm and value. In 
	{\em Proceedings of the ASL Meeting 1990, Helsinki}, 
	Juha Oikkonen and Jouko V\"a\"an\"anen (eds.), 210 -- 249.
	(Lecture Notes in Logic 2.) Berlin and Heidelberg: Springer.
\end{eintrag}
\\[2.7mm]
Myhill, John
\\\begin{eintrag}
1960 & Linear bounded automata. Technical report, Wright--Patterson 
	Air Force Base.
\end{eintrag}
\\[2.7mm]
Ogden, R.~W., Ross, R,~J., and Winkelmann, K.
\\\begin{eintrag}
1985 & An ``{I}nterchange {L}emma'' for {C}ontext {F}ree {L}anguages.
	{\em SIAM Journal of Computing} 14:410 -- 415.
\end{eintrag}
\\[2.7mm]
Ogden, R.~W.
\\\begin{eintrag}
1968 & A helpful result for proving inherent ambiguity. {\em Mathematical 
	Systems Sciences} 2:191 -- 194.
\end{eintrag}
\\[2.7mm]
Ojeda, Almerindo~E. 
\\\begin{eintrag}
1988 & A {L}inear {P}recedence {A}ccount of {C}ross--{S}erial
  {D}ependencies. {\em Linguistics and Philosophy} 11:457 -- 492.
\end{eintrag}
\\[2.7mm]
Pentus, Mati 
\\\begin{eintrag}
1995 & Models for the {L}ambek calculus. {\em Annals of Pure and 
	Applied Logic} 75:179 -- 213.
\\
1997 & {P}roduct--{F}ree {L}ambek--{C}alculus and {C}ontext--{F}ree
  {G}rammars. {\em Journal of Symbolic Logic} 62:648 -- 660.
\end{eintrag}
\\[2.7mm]
Peters, Stanley~P., and Ritchie, R.~W.
\\\begin{eintrag}
1971 & On restricting the base component of transformational grammars.
	{\em Information and Control} 18:483 -- 501.
\\
1973 & On the generative power of transformational grammars.
	{\em Information Sciences} 6:49 -- 83.
\end{eintrag}
\\[2.7mm]
Pigozzi, Don J., and Salibra, Antonino
\\\begin{eintrag}
1995 & The {A}bstract {V}ariable--{B}inding {C}alculus.
	{\em Studia Logica} 55:129 -- 179.
\end{eintrag}
\\[2.7mm]
Pigozzi, Don~J. 
\\\begin{eintrag}
1991 & Fregean {A}lgebraic {L}ogic. In {\em Algebraic Logic}, 
	Hajnal Andr\'eka, Donald Monk, and Istv\'an N\'emeti (eds.), 
	475 -- 504. (Colloquia Mathematica Societatis J\'anos Bolyai 54.), 
	Budapest and Amsterdam: J\'anos Bolyai Matematikai 
	T\'arsulat and North--Holland.
\end{eintrag}
\\[2.7mm]
Pogodalla, Sylvain
\\\begin{eintrag}
2001 & {\em R\'eseaux de preuve et g\'en\'eration pour les grammaires de
  type logiques} [Proof nets and generation for type logical grammars].
	Ph.\ D.\ diss., Institut National Polytechnique de Lorraine.
\end{eintrag}
\newpage\noindent
Pollard, Carl J., and Sag, Ivan 
\\\begin{eintrag}
1987 & {\em Information--{B}ased {S}yntax and {S}emantics. Vol.~1}.
	CSLI Lecture Notes 13. Stanford: CSLI.
\\
1994 & {\em Head--Driven Phrase Structure Grammar}. Chicago: 
	The University of Chicago Press.
\end{eintrag}
\\[2.2mm]
Pollard, Carl~J. 
\\\begin{eintrag}
1984 & {\em Generalized Phrase Structure Grammar, Head Grammars and 
	Natural Language}. Ph.\ D.\ diss., Stanford University.
\end{eintrag}
\\[2.2mm]
Port, R.~F., and O'Dell, M.~L. 
\\\begin{eintrag}
1985 & Neutralization of syllable--final voicing in {G}erman.
	Technical Report~4, Indiana University, Bloomington.
\end{eintrag}
\\[2.2mm]
Post, Emil~L. 
\\\begin{eintrag}
1936 & Finite combinatory processes -- formulation. {\em Journal of 
	Symbolic Logic} 1:103 -- 105.
\\
1943 & Formal reductions of the combinatorial decision problem.
	{\em Americal Journal of Mathematics} 65:197 -- 215.
\\
1947 & Recursive unsolvability of a problem of {T}hue.
	{\em Journal of Symbolic Logic} 11:1 -- 11.
\end{eintrag}
\\[2.2mm]
Postal, Paul 
\\\begin{eintrag}
1964 & {\em {C}onstituent {S}tructure: {A} {S}tudy of {C}ontemporary
  {M}odels of {S}yntax}. The Hague: Mouton.
\end{eintrag}
\\[2.2mm]
Prucnal, T., and Wro\'{n}ski, A.
\\\begin{eintrag}
1974 & An algebraic characterization of the notion of structural
  completeness. {\em Bulletin of Section Logic of the Polish Academy 
	of Sciences} 3:20 -- 33.
\end{eintrag}
\\[2.2mm]
Quine, Willard van Orman 
\\\begin{eintrag}
1960 & Variables explained away. {\em Proceedings of American 
	Philosophical Society} 104:343 -- 347.
\end{eintrag}
\\[2.2mm]
Radzinski, Daniel
\\\begin{eintrag}
1990 & {U}nbounded {S}yntactic {C}opying in {M}andarin {C}hinese.
	{\em Linguistics and Philosophy} 13:113 -- 127, 1990.
\end{eintrag}
\\[2.2mm]
Rambow, Owen 
\\\begin{eintrag}
1994 & {\em Formal and Computational Aspects of Natural Language Syntax}.
	Ph.\ D.\ diss., University of Pennsylvania.
\end{eintrag}
\\[2.2mm]
Recanati, Fran\c{c}ois 
\\\begin{eintrag}
2000 & {\em Oratio Obliqua, Oratio Recta. An Essay on Metarepresentation}.
	Cambridge (Mass.): MIT Press.
\end{eintrag}
\\[2.2mm]
Roach, Kelly
\\\begin{eintrag}
1987 & Formal properties of head grammars. In {\em Mathematics of Language}, 
	Alexis Manaster{--}Ramer (ed.), 293--347. 
	Amsterdam: John Benjamins.
\end{eintrag}
\newpage\noindent
Rogers, James 
\\\begin{eintrag}
1994 & {\em Studies in the {L}ogic of {T}rees with {A}pplications to
  {G}rammar {F}ormalisms}. Ph.\ D.\ diss., University of Delaware, 
  Department of Computer \& Information Sciences.
\end{eintrag}
\\[2.5mm]
Rounds, William~C. 
\\\begin{eintrag}
1988 & {LFP}: {A} {L}ogic for {L}inguistic {D}escription and an {A}nalysis
  of its {C}omplexity. {\em Computational Linguistics} 14:1 -- 9.
\end{eintrag}
\\[2.5mm]
Russell, Bertrand 
\\\begin{eintrag}
1905 & On denoting. {\em Mind} 14:479 -- 493.
\end{eintrag}
\\[2.5mm]
Sain, Ildik\'{o}, and Thompson, Richard~S. 
\\\begin{eintrag}
1991 & Finite {S}chema {A}xiomatization of {Q}uasi--{P}olyadic {A}lgebras.
	In {\em Algebraic Logic}, Hajnal Andr\'eka, Donald Monk, and 
	Istv\'an N\'emeti (eds.), 539 -- 571. (Colloquia Mathematica 
	Societatis J\'anos Bolyai 54.) Budapest and Amsterdam: J\'anos 
	Bolyai Matematikai T\'arsulat and North--Holland.
\end{eintrag}
\\[2.5mm]
Salomaa, Arto~K. 
\\\begin{eintrag}
1973 & {\em Formal Languages}. New York: Academic Press.
\end{eintrag}
\\[2.5mm]
van~der Sandt, Rob A.
\\\begin{eintrag}
1988 & {\em Context and Presupposition}. London: Croom Helm.
\end{eintrag}
\\[2.5mm]
de Saussure, Ferdinand
\\\begin{eintrag}
1965 & {\em Course in General Linguistics}. Columbus: McGraw--Hill.
\end{eintrag}
\\[2.5mm]
Sauvageot, Aur\'{e}lien 
\\\begin{eintrag}
1971 & {\em L'{E}dification de la {L}angue {H}ongroise\/} [The building of
  the Hungarian language]. Paris: Klincksieck.
\end{eintrag}
\\[2.5mm]
Sch\"onfinkel, Moses
\\\begin{eintrag}
1924 & {\"U}ber die {B}austeine der ma\-the\-ma\-ti\-schen {L}ogik [{O}n
  the building blocks of mathematical logic]. {\em Mathematische Annalen} 
	92:305 -- 316.
\end{eintrag}
\\[2.5mm]
Seki, Hiroyuki, Matsumura, Takashi, Fujii, Mamoru, and Kasami, Tadao.
\\\begin{eintrag}
1991 & On multiple context--free grammars. {\em Theoretical Computer 
	Science} 88:191 -- 229.
\end{eintrag}
\\[2.5mm]
Sestier, A.
\\\begin{eintrag}
1960 & Contributions \`{a} une th\'{e}orie ensembliste des classifications
  linguistiques [{C}ontributions to a set--theoretical theory of
  classifications]. In {\em Actes du Ier Congr\`{e}s de l'AFCAL}, 
  293 -- 305, Grenoble.
\end{eintrag}
\\[2.5mm]
Shieber, Stuart 
\\\begin{eintrag}
1985 & Evidence against the {C}ontext--{F}reeness of {N}atural 
	{L}anguages. {\em Linguistics and Philosophy} 8:333 -- 343.
\end{eintrag}
\\
\begin{eintrag}
1992 & {\em Constraint--{B}ased {G}rammar {F}ormalisms}. 
	Cambridge (Mass.): MIT Press.
\end{eintrag}
\\[3.3mm]
Smullyan, Raymond~M. 
\\\begin{eintrag}
1961 & {\em Theory of Formal Systems}. Annals of Mathematics 
	Studies 47. Princeton: Princeton University Press.
\end{eintrag}
\\[3.3mm]
Staal, J.~F. 
\\\begin{eintrag}
1967 & {\em Word Order in Sanskrit and Universal Grammar}.
	Foundations of Language, Supplementary Series No.~5. 
	Dordrecht: Reidel.
\end{eintrag}
\\[3.3mm]
Stabler, Edward~P. 
\\\begin{eintrag}
1997 & Derivational {M}inimalism. In {\em Logical Aspects of Computational 
	Linguistics (LACL '96)}, Christian Retor\'{e} (ed.),  68 -- 95.
	(Lecture Notes in Artificial Intelligence 1328.)
	Heidelberg: Springer.
\end{eintrag}
\\[3.3mm]
von Stechow, Arnim, and Sternefeld, Wolfgang
\\\begin{eintrag}
1987 & {\em Bausteine syntaktischen Wissens. Ein Lehrbuch der generativen
  Grammatik\/} [{B}uilding blocks of syntactic knowledge. A textbook of
  generative grammar]. Opladen: Westdeutscher Verlag.
\end{eintrag}
\\[3.3mm]
Steedman, Mark 
\\\begin{eintrag}
1990 & Gapping as constituent coordination. {\em Linguistics and 
	Philosophy} 13:207 -- 263.
\\
1996 & {\em Surface Structure and Interpretation}. 
	Linguistic Inquiry Monographs 30. Cambridge (Mass.): 
	MIT Press.
\end{eintrag}
\\[3.3mm]
Tarski, Alfred 
\\\begin{eintrag}
1983 & The concept of truth in formalized languages. In 
	{\em Logic, Semantics, Metamathematics}, J.~Corcoran (ed.), 
	152 -- 178. Indianapolis: Hackett Publishing.
\end{eintrag}
\\[3.3mm]
Tesni\`ere, Lucien
\\\begin{eintrag}
1982 & {\em El\'ements de syntaxe structurale\/} [Elements of structural
  syntax]. 4th ed. Paris: Klincksieck.
\end{eintrag}
\\[3.3mm]
Thatcher, J.~W., and Wright, J.~B.
\\\begin{eintrag}
1968 & Generalized finite automata theory with an application to a 
	decision problem of second--order logic. {\em Mathematical 
	Systems Theory} 2:57 -- 81.
\end{eintrag}
\\[3.3mm]
Thue, Axel 
\\\begin{eintrag}
1914 & Probleme \"uber {V}er\"anderungen von {Z}ei\-chen\-rei\-hen nach
  gegebenen {R}egeln [{P}roblems concerning changing strings according to
  given rules]. {\em Skrifter utgit av Videnskapsselkapet i Kristiania, {I}.
  Mathematisk--naturvidenskabelig klasse} 10.
\end{eintrag}
\newpage
\noindent
Trakht\'enbrodt, B.~A.
\\\begin{eintrag}
1950 & N\'evozmo\v{z}nost' algorifma dl\'a probl\'emy razr\'e\v{s}imosti na
  kon\'e\v{c}nyh klassah [{I}mpossibility of an algorithm for the decision
  problem of finite classes]. {\em Doklady Akad\'emii Nauk SSSR}, 
  569 -- 572.
\end{eintrag}
\\[2.3mm]
Turing, Alan~M. 
\\\begin{eintrag}
1936 & On computable numbers, with an application to the 
	{E}ntscheidungsproblem. {\em Proceedings of the London 
	Mathematical Society} 42:230 -- 265.
\end{eintrag}
\\[2.3mm]
Uszkoreit, Hans
\\\begin{eintrag}
1987 & {\em Word {O}rder and {C}onstituent {S}tructure in {G}erman}.
	CSLI Lecture Notes 8. Stanford: CSLI.
\end{eintrag}
\\[2.3mm]
Vaught, Robert~L. 
\\\begin{eintrag}
1995 & {\em Set Theory. An Introduction}. 2nd ed. Basel: 
	Birkh\"auser.
\end{eintrag}
\\[2.3mm]
Frank Veltman.
\\\begin{eintrag}
1985 & {\em Logics for Conditionals}. Ph.\ D.\ diss., Department of 
	Philosophy, University of Amsterdam.
\end{eintrag}
\\[2.3mm]
Vijay{--}Shanker, K., Weir, David~J., and Joshi, Aravind~K. 
\\\begin{eintrag}
1986 & Tree adjoining and head wrapping. In {\em Proceedings of the 11th 
	International Conference on {C}omputational {L}inguistics 
	({COLING} '86)\emph{, Bonn}}, 202--207.
\\
1987 & Characterizing structural descriptions produced by various grammar
	  formalisms. In {\em Proceedings of the 25th Meeting of the 
	Association for Computational Linguistics ({ACL} '87)\emph{, 
	Stanford, CA}}, 104 -- 111.
\end{eintrag}
\\[2.3mm]
Villemonte de~la Clergerie, Eric
\\\begin{eintrag}
2002a & Parsing {MCS} {L}anguages with {T}hread {A}utomata.
	In {\em Proceedings of the Sixth International Workshop on Tree
  Adjoining Grammars and Related Formalisms (TAG+6), \emph{Venezia}}.
\\
2002b & Parsing mildly context--sensitive languages with thread automata.
	In {\em Proceedings of the 19th International Conference on
        Computational Linguistics (COLING 02)\emph{, Taipei}}.
\end{eintrag}
\\[2.3mm]
Weir, David~J. 
\\\begin{eintrag}
1988 & {\em Characterizing {M}ildly {C}ontext--{S}ensitive {G}rammar
  {F}ormalisms}. Ph.\ D.\ diss., University of Pennsylvania, Philadelphia.
\end{eintrag}
\\[2.3mm]
Younger, D.~H. 
\\\begin{eintrag}
1967 & Recognition and parsing of context--free languages in time $n^3$.
	{\em Information and Control} 10:189 -- 208.
\end{eintrag}
\\[2.3mm]
Zadrozny, Wlodek
\\\begin{eintrag}
1994 & From {C}ompositional {S}emantics to {S}ystematic {S}emantics.
	{\em Linguistics and Philosophy} 17:329 -- 342.
\end{eintrag}
\\[2mm]
Zimmermann, Thomas~Ede 
\\\begin{eintrag}
1999 & Meaning {P}ostulates and the {M}odel--{T}heoretic {A}pproach to
  {N}atural {L}anguage {S}emantics. {\em Linguistics and Philosophy} 
	22:529 -- 561.
\end{eintrag}
